Age-structured transmission dynamic models of RSV

### 2023-01-02 MSIS_model_explain/MSIS_model_explain.Rmd
add stan codes section

```{r,echo=T,eval=FALSE}
fitLL <- optim(par=c(2.2,-1.6,1.2,3.16),
  # remember to exponential these values
  # to get the parameter estimates
  # starting values for beta and gamma 
  # - you should get the similar/same result 
  # no matter which values you choose here
      fn = fitmodel, # the distance function to optimise
      dat = sim,  # the dataset we fit to 
# ("dat" argument is passed to the function specified in fn)
    control = list(fnscale=-1))# negative log likelihood
# here we minimize the negative log likelihood
```

@@

exp(fitLL$par)
# output: 8.99  0.20  3.3 (same as+/-pi) 
# 23

# Model input:
# 9, 0.2, 3.33 30 days

@@

Before we start, please learn the [Rstan basic](https://github.com/Gigi112/RSV_tutorials/tree/main/MSIS_model_explain/Introduction of Rstan.pptx) from the seperated tutorials in the same file path.

>>>>

```{r,echo=T,eval=FALSE}
fitLL <- optim(par=c(2.2,-1.6,0.1,3.5),
  # remember to exponential these values
  # to get the parameter estimates
  # starting values for beta and gamma 
  # - you should get the similar/same result 
  # no matter which values you choose here
      fn = fitmodel, # the distance function to optimise
      dat = sim,  # the dataset we fit to 
# ("dat" argument is passed to the function specified in fn)
    control = list(fnscale=-1))# negative log likelihood
# here we minimize the negative log likelihood
```

@@

exp(fitLL$par)[c(1:2)] # trans prob and Amp est
phi_est = (2*pi*(exp(fitLL$par[3]))) / (1+exp(fitLL$par[3]))
dur_est=exp(fitLL$par)[4]
# output: 8.99  0.20  
# 3.33 (same as+/-pi) 
# 23.4 days

# Model input:
# 9, 0.2, 3.33, 30 days

@@

Before we start, please learn the [Rstan basic](https://github.com/Gigi112/RSV_tutorials/tree/main/MSIS_model_explain/Introduction of Rstan.pptx) from the separated tutorials in the same file path. 
\
\
You will find three code documents inside code_source/stan_code file. age-structureMSIS_deterministic.stan is the stan code contains the ODE function, input data, input parameters, priors and the distribution of posterior distributions. ageMSIS_deterministric_annual.R is the R code that create the interface between stan and R to sample from posterior distribution. msis_wa_D.sh is a file to run the model on cluster. We will start with the stan code.
\
\
I suggest you first run the model in local for short iterations for debug purpose. If the stan model works, you can move to HPC by using the msis_wa_D.sh script.

## Stan code in details (page 55 to 68)
Function: page 55 to 60.
```{stan,output.var="model",echo=T,eval=FALSE}
// function is the similar as our simpleModel.R
// except for the syntax.
functions {
  real[] msis (real t, real[] y, real[] theta,
             real[] x_r, int[] x_i) {
// t for time

// state, the volumes in each compartment, y;

// theta, variables used to compute function
// which depend on the model parameters;

// x_r, real variables used to evaluate f
// which only depend on fixed data;

// x_i, integer values used to evaluate f
// which only depend on fixed data.

      int agegroups = x_i[1];
      int q=x_i[2]; // Freq or Density dependent  
```

## You need to specify the variable type and locations of the parameter for each value
Note, remember to add **;** after each line of code.
\
```{stan,output.var="model",echo=T,eval=FALSE}
real birthrate[agegroups] = x_r[1:agegroups];
real um = x_r[agegroups+1]; // net growth
real rho1 = x_r[agegroups+2];
real rho2 = x_r[agegroups+3];
real gamma1 = x_r[agegroups+4];
real gamma2 = x_r[agegroups+5];
real gamma3 = x_r[agegroups+6];
real sigma1 = x_r[agegroups+7];
real sigma2 = x_r[agegroups+8];
real sigma3 = x_r[agegroups+9];
real u[agegroups]= x_r[(agegroups+10):(9+2*agegroups)];
//aging rate
real c2[agegroups*agegroups]=
      x_r[(10+2*agegroups):(9+2*agegroups+agegroups*agegroups)];
      // contact vector 
// (I have not yet figured out how to input matrix)
```

## When create a new variable, you also need to specify the length of the variable 
```{stan,output.var="model",echo=T,eval=FALSE}
// parameters to estimate
real beta = theta[1];
real b1 = theta[2];
real phi = theta[3];
real omega = theta[4];

// outcomes of interests that need to be recorded      
real dydt[9*agegroups]; // state change
real lambda[agegroups]; // force of infection
real InfectN[agegroups];// number of infectious individuals

// seasonal component
real season_txn = (1+b1*cos(2*pi()*(t-phi*12)/12));
```

## Write down ordinary differential equations
```{stan,output.var="model",echo=T,eval=FALSE}
for (k in 1:agegroups) {
      InfectN[k] = (y[k+2*agegroups]+
      rho1*y[k+4*agegroups]+
      rho2*y[k+6*agegroups]+
      rho2*y[k+8*agegroups]);
      } // (I1+rho1*I2+rho2*I3+rho2*I4)

for (a in 1:agegroups) {
  // force of infection    
  lambda[a] = season_txn*beta*gamma1/sum(y)*
  sum(to_vector(c2[((a-1)*agegroups+1):(a*agegroups)])
  .*to_vector(InfectN));

   //correspond to M.     
  dydt[a] =  log(birthrate[a]+1)/12*sum(y) - 
  (omega+u[a]+um)*y[a];
```

## Ordinary differential equations (continued)
corresponding R code: [Source function of the transmission dynamic model (continued)] except for the aging in process (include in the next page).
```{stan,output.var="model",echo=T,eval=FALSE}
//S0
  dydt[a+agegroups] =  omega*y[a] - 
  lambda[a]*y[a+agegroups] - 
  (um+u[a])*y[a+agegroups];

//I1  
 dydt[a+2*agegroups] =  lambda[a]*y[a+agegroups] - 
 gamma1*y[a+2*agegroups] - 
 (um+u[a])*y[a+2*agegroups];

//S1 
dydt[a+3*agegroups] =  gamma1*y[a+2*agegroups] - 
sigma1*lambda[a]*y[a+3*agegroups] - 
(um+u[a])*y[a+3*agegroups];
```

We will skip explaining the rest of the stan code of ordinary differential equations in this tutorial as they are the direct translation of the corresponding R codes.

## Aging in process
In age group >= 1 month, they receive the inflows from the younger groups at rate u. 
```{stan,output.var="model",echo=T,eval=FALSE}
if (a >1 ){
dydt[a] = dydt[a]+ u[a-1]*y[a-1];

dydt[a+agegroups] = dydt[a+agegroups]+ 
u[a-1]*y[a+agegroups-1];

dydt[a+2*agegroups] = dydt[a+2*agegroups] + 
u[a-1]*y[a+2*agegroups-1];

//...//

dydt[a+8*agegroups] = dydt[a+8*agegroups]+
u[a-1]*y[a+8*agegroups-1];}

 }
    return dydt;
  }
} // This is the end of the ODE function
```

## Data block

Compared to R, the data block in stan code contains both fixed parameter values and also input observed data.

```{stan,output.var="model",echo=T,eval=FALSE}
// Fixed data is declared in the data block:
data {
  int<lower=1> n_months;
  int<lower=1> agegroups;
  int q;
  real y0[9*agegroups];
  real t0;
  real ts[n_months];
  int N;
  real birthrate[agegroups];
  real um;
//...omit some parameters and data to fit into slides//
  int<lower=1> hosp_cases[n_months];
  int hosp_age[8];
  real hosp1[agegroups];
  real c2[agegroups*agegroups];
}
```

## Transformed data block
Link data block to function inputs: real[ ] x_r, int[ ] x_i
```{stan,output.var="model",echo=T,eval=FALSE}
transformed data {
  real x_r[204];
  int x_i[2];
  x_i[1] = agegroups;
// the first value of x_i is the number of age groups
  x_i[2] = q;
// the second value of x_i is parameter q

  x_r[1:agegroups]= birthrate;
// the first value of x_r is birth rate
  x_r[agegroups+1]=um;
//likewise, omit some parameters and data to fit into slides//
  x_r[agegroups+9]=sigma3;
  x_r[(agegroups+10):(9+2*agegroups)]=u;
  x_r[(10+2*agegroups):(9+2*agegroups+agegroups*agegroups)]=c2;
}
```

## Parameter block
Specify the parameters we would like to estimate and their ranges.
```{stan,output.var="model",echo=T,eval=FALSE}
parameters {
  real<lower=0> beta; // transmission prob
  real<lower=0,upper=1> b1; // Amp
  real<lower=0,upper=2*pi()> phi; // phase
  real<lower=0> omega;
  // waning rate of maternal immunity
  real<lower=0,upper=1> report_ratio;
  // report ratio of RSV hospitalizations
  }
```

## Transformed parameters clock

This part of stan codes corresponds to using ode() function in r to get the transmission states of each age group in each time step and then calculate the number of RSV hospitalizations and age distribution.
```{stan,output.var="model",echo=T,eval=FALSE}
transformed parameters{
real y[n_months, 9*agegroups]; 
// y is the states matrix that has row length=n_months and columns=9.
  real rel_inf[n_months,agegroups];
// number of infectious individuals
  real lambda[n_months, agegroups]; // force of infection
  // outcomes
  // Number of hospitalizations by age
  matrix[n_months,agegroups] HOSP_AGE;// corresponds to H1
  vector[8] output_hosp_age; 
  // age distribution of RSV hosp in children
  vector[n_months] total_children; 
  // total hospitalizations in children <5 years
  vector<lower = 0>[n_months] output_hosp; 
  //number of RSV hospitalizations by month
```

## Transformed parameters clock (continued)
```{stan,output.var="model",echo=T,eval=FALSE}
  real theta[4]; // parameters that need to be estimate
  {
    theta[1] = beta;
    theta[2] = b1;
    theta[3] = phi;
    theta[4] = omega;

    y = integrate_ode_bdf(msis, y0, t0, ts, theta, x_r, x_i,
    1.0E-10, 1.0E-10, 1.0E3);
    // relative tolerance for the ODE solver = 1.0E-10
    // absolute tolerance for the ODE solver = 1.0E-10
    // maximum number of steps to take in the ODE solver
    // = 1.0E3
    // these prevent stan takes too long to run
    }
```

## Calculate force of infection (continued transformed parameters)

```{stan,output.var="model",echo=T,eval=FALSE}
for(i in 1:n_months){
  for(j in 1:9*agegroups){
    if (y[i,j] <= 0.0) y[i,j] = 1e-12;}
// this prevent negative values in states

      for (k in 1:agegroups) {
      rel_inf[i,k] = (y[i,k+2*agegroups]+
      rho1*y[i,k+4*agegroups]+
      rho2*y[i,k+6*agegroups]+
      rho2*y[i,k+8*agegroups]);
      }// (I1+rho1*I2+rho2*I3+rho2*I4)

// calculate force of infection; same as inside function     
      for (a in 1:agegroups) {
       lambda[i,a] = (1+b1*cos(2*pi()*(i-phi*12)/12))*
       beta*gamma1/sum(y[i,])*
       sum(to_vector(c2[((a-1)*agegroups+1):(a*agegroups)])
       .*to_vector(rel_inf[i,]));
```

## Calculate hospitalizations (continued transformed parameters)

```{stan,output.var="model",echo=T,eval=FALSE}
HOSP_AGE[i,a] = hosp1[a]*lambda[i,a]*y[i,a+agegroups]
			+hosp2[a]*lambda[i,a]*sigma1*y[i,a+3*agegroups]	
			+hosp3[a]*lambda[i,a]*sigma2*y[i,a+5*agegroups]
			+hosp3[a]*lambda[i,a]*sigma3*y[i,a+7*agegroups];
       } // same as H1 calculation

  output_hosp[i] = report_ratio*sum(HOSP_AGE[i,]);
  // total reported RSV hospitalizations each month

  total_children[i]=HOSP_AGE[i,1]+HOSP_AGE[i,2]+
  HOSP_AGE[i,3]+HOSP_AGE[i,4]+HOSP_AGE[i,5]+HOSP_AGE[i,6]+
  HOSP_AGE[i,7]+HOSP_AGE[i,8]; 
  // total hospitalizations in children <5 years
    }
for (b in 1:8) {
 output_hosp_age[b] = sum(HOSP_AGE[,b])/sum(total_children);}} 
 // age distribution of hospitalizations in children <5 years
```

## Model block

(1) specify the priors of the parameters we would like to estimate and their ranges.
/
/
(2) specify the sample distributions of the posteriors.
```{stan,output.var="model",echo=T,eval=FALSE}

model {
  //priors
  beta ~ normal(3,1) T[1,5]; //truncated at 0
  b1 ~ normal(0.2,0.05) T[0.05,1]; //truncated at 0,1
  phi ~ normal(1.4,0.5) T[0,2*pi()];
  omega ~ lognormal(-1,0.6) T[0,5];
  report_ratio ~ beta(2,2); //truncated at 0,1

  //sampling distribution
  for(i in 1:n_months){
  target += poisson_lpmf(hosp_cases[i]| output_hosp[i]);}

  target += multinomial_lpmf (hosp_age|output_hosp_age);
}
```

## R function to call stan code

```{r,echo=T,eval=FALSE}
library(rstan) # function to link stan code and R code
rstan_options (auto_write = TRUE) 
# this code is recommended for running stan in HPC

setwd("~/project/Rstan/WA") 
# this is where I save the data and all these codes
data_msis<- readRDS("data_msis_wa_annual.rds")
# read in input data. 
# This would come from the inpatient data sets
# and known parameters

print(paste("number of cores:", future::availableCores()))
# this is to check the cores you can use in HPC/local

# call the stan code
model <- stan_model("age-structureMSIS_deterministic.stan")

# set the initial values as the output of MLE function
# This helps the running time
init_fun <- function() list(beta=3.17,b1=0.198,
                    phi=1.46,omega=0.61,hosp_ratio=0.76)
```

## R function to sample the posteriors from stan code

```{r,echo=T,eval=FALSE}
## this code samples the posteriors of parameters of interests
fit_msis <- sampling(model, # the stan model
            data = data_msis, 
          # inpatient data and known parameters
            iter = 2000,# total iterations
			      warmup = 1000, #iterations discard
            chains = 4, # total chains
            seed = 112, # for reproducible purpose
            sample_file = file.path(getwd(), 
              'WA_MSIS_12_deterministic.csv'),
          # this is where the posterior samples are saved
            cores=4,#parallel: one core for one chain
            open_progress=T, # to see immediate output
			      init= init_fun, # initial values
            control = list(stepsize=0.1)) # sample step size 
```

## Shell script (bash codes, Sample SLURM Scripts) to run R and STAN on high performance computer (HPC)
```{r,echo=T, engine = 'bash',eval=FALSE}
# indicate the script uses the bash shell.
#!/bin/bash

#SBATCH --time=30-  
# This tells the cluster the running time for your script. 
# Don't set it too high or your job will get low priority. 
#SBATCH --mail-type=ALL 
# This tells the cluster what results should send to your email.
#SBATCH --mail-user=zhe.zheng@yale.edu 
# This is your email address
#SBATCH --cpus-per-task=4 
# This is for running in parallel. 
# Set the cores to the number of chains you have
#SBATCH --mem-per-cpu=30000 
# This is for memory for each core

module load R/4.1.0-foss-2020b 
# you first need to check the r version available on cluster
Rscript ageMSIS_deterministric_annual.R # R script to run
```
